---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# taxastand

<!-- badges: start -->
<!-- badges: end -->

The goal of `taxastand` is to standardize species names from different sources, a common task in biology. Very often different biologists use different synonyms to refer to the same species. If we want to join data from different sources, their taxonomic names must be standardized first. This is what `taxastand` seeks to do in a reproducible and efficient manner.

`taxastand` is based on matching names to a single **taxonomic standard**, that is, a database of accepted names and synonyms. As long as a single taxonomic standard is used, we can confidently resolve names from disparate sources.

The taxonomic standard must conform to [Darwin Core standards](https://dwc.tdwg.org/). The user must provide this database (as a dataframe). There are many sources of taxonomic data online, including [GBIF](https://www.gbif.org/en/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c), [Catalog of Life](http://www.catalogueoflife.org/), and [ITIS](https://www.itis.gov/) to name a few. The [taxadb](https://cboettig.github.io/taxadb/index.html) package provides convenient functions for downloading various taxonomic databases that use Darwin Core.

## Installation

`taxastand` is currently only available on GitHub:

``` r
install.packages("devtools")
devtools::install_github("joelnitta/taxastand")
```

## Similar work

[ROpenSci](https://ropensci.org/) has a [task view](https://github.com/ropensci/taxonomy) summarizing many tools available for taxonomy.

[taxize](https://github.com/ropensci/taxize) is the "granddaddy" of taxonomy packages in R. It can search around 20 different taxonomic databases for names and retrieve taxonomic information. It does not store any databases locally, and does not do fuzzy matching.

[taxizedb](https://github.com/ropensci/taxizedb) downloads taxonomic databases and provides tools to interface with them through SQL.

[taxadb](https://cboettig.github.io/taxadb/index.html) also downloads and searches taxonomic databases. It can interface with them either through SQL or in-memory in R.

## Examples

Here are some examples that run quickly using the small example dataset included with the package.

```{r filmy-example}
library(taxastand)

# Load reference taxonomy in Darwin Core format
data(filmy_taxonomy)

# This taxon matches many names at the species level because
# there are a bunch of varieties.
match_taxonomy("Hymenophyllum polyanthos", filmy_taxonomy, "species")

# Using the full species name with author can get us a
# more exact match.
match_taxonomy("Hymenophyllum polyanthos (Sw.) Sw.", filmy_taxonomy, "sciname")

# Fuzzy match helps when the query didn't abbreviate
# the author, but it is abbreviated in the reference.
match_taxonomy("Hymenophyllum polyanthos (Swartz) Swartz",
filmy_taxonomy, "sciname", max_dist = 8)
```

Here is a more "real-world" type of example that involves downloading the entire GBIF database using `taxadb`.

(This is more realistic, but is too slow right now. Needs to be optimized.)

```{r taxadb-example}
library(taxadb)
library(tictoc)

# The taxa_tbl() function downloads a local copy of a taxonomic database from various providers.
gbif <- taxa_tbl("gbif")

tic() # set a timer
match_taxonomy("Hymenophyllum polyanthos", gbif, "species")
toc() # see how long it took
```
