---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# taxastand

<!-- badges: start -->
<!-- badges: end -->

**DISCLAIMER:** This package is under **active development**. Functions may change drastically, and the user should not expect the interface to stay consistent. Many functions are not yet available.

The goal of `taxastand` is to standardize species names from different sources, a common task in biology. Very often different biologists use different synonyms to refer to the same species. If we want to join data from different sources, their taxonomic names must be standardized first. This is what `taxastand` seeks to do in a reproducible and efficient manner.

`taxastand` is based on matching names to a single **taxonomic standard**, that is, a database of accepted names and synonyms. As long as a single taxonomic standard is used, we can confidently resolve names from disparate sources.

The taxonomic standard must conform to [Darwin Core standards](https://dwc.tdwg.org/). The user must provide this database (as a dataframe). There are many sources of taxonomic data online, including [GBIF](https://www.gbif.org/en/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c), [Catalog of Life](http://www.catalogueoflife.org/), and [ITIS](https://www.itis.gov/) to name a few. The [taxadb](https://cboettig.github.io/taxadb/index.html) package provides convenient functions for downloading various taxonomic databases that use Darwin Core.

## Installation

`taxastand` is currently only available on GitHub:

``` r
install.packages("devtools")

# Install another github-hosted dependency first
devtools::install_github("joelnitta/jntools")

devtools::install_github("joelnitta/taxastand")
```

## Similar work

[ROpenSci](https://ropensci.org/) has a [task view](https://github.com/ropensci/taxonomy) summarizing many tools available for taxonomy.

[taxize](https://github.com/ropensci/taxize) is the "granddaddy" of taxonomy packages in R. It can search around 20 different taxonomic databases for names and retrieve taxonomic information.

[TNRS](http://tnrs.iplantcollaborative.org/), the Taxonomic Name Resolution Service, is a web application that resolves taxonomic names of plants according to one of six databases.

[taxizedb](https://github.com/ropensci/taxizedb) downloads taxonomic databases and provides tools to interface with them through SQL.

[taxadb](https://cboettig.github.io/taxadb/index.html) also downloads and searches taxonomic databases. It can interface with them either through SQL or in-memory in R.

[taxonstand](https://cran.r-project.org/web/packages/Taxonstand/index.html) has a very similar goal to `taxastand`, but only uses [The Plant List](www.theplantlist.org) as its taxonomic standard and does not allow the user to provide their own.

## Motivation

Although existing web-based solutions for taxonomic name resolution are very useful, they suffer some drawbacks: the choice of database to use for standardization is limited, they may not be appropriate for very large queries, and the user has no guarantee that the same input will yield the same output at a later date due to changes in code or database version.

The motivation for `taxastand` is to provide greater flexibility and reproducibility by allowing for complete version control of the code and database used for name resolution.

## Examples

Here are some name-matching examples that run quickly using the small example dataset included with the package.

```{r filmy-example}
library(taxastand)

# Load reference taxonomy in Darwin Core format
data(filmy_taxonomy)

# This taxon matches many names at the species level because
# there are a bunch of varieties.
match_taxonomy("Hymenophyllum polyanthos", filmy_taxonomy, "species")

# Using the full species name with author can get us a
# more exact match.
match_taxonomy("Hymenophyllum polyanthos (Sw.) Sw.", filmy_taxonomy, "scientific_name")

# Fuzzy match helps when the query didn't abbreviate
# the author, but it is abbreviated in the reference.
match_taxonomy("Hymenophyllum polyanthos (Swartz) Swartz",
filmy_taxonomy, "scientific_name", max_dist = 8)
```
