---
title: "Benchmark: pure R vs. taxon-tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmark: pure R vs. taxon-tools}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r detect, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# The benchmark compares against the original taxon-tools, run from its Docker
# image. The image bundles the aregex (gawk amatch) build of taxon-tools, i.e.
# the faster of the tool's fuzzy-matching backends. The benchmark chunks below
# are only evaluated when Docker and that image are available, so this vignette
# still builds (showing a note instead of tables) on machines without Docker.
image <- "camwebb/taxon-tools:v1.3.0"
has_docker <- tryCatch(
  system2("docker", c("image", "inspect", image),
    stdout = FALSE, stderr = FALSE
  ) == 0,
  error = function(e) FALSE
)
```

## Background

As of version 2.0.0, `taxastand` parses and matches taxonomic names with a
pure-R reimplementation of [taxon-tools](https://github.com/camwebb/taxon-tools)
(`parsenames` and `matchnames`). Earlier versions shelled out to the original
`taxon-tools` programs, either via a local install or the
`r image` Docker image.

The pure-R engines produce **byte-for-byte identical output** to the original
tool (this is checked in the package's test suite against the Docker image). The
question this vignette answers is: how do they compare on **speed**?

The comparison below is against the `aregex` build of `taxon-tools` shipped in
the Docker image, which uses gawk's `amatch()` for fuzzy matching --- the faster
of the tool's two fuzzy backends.

## Setup

```{r setup, eval = has_docker}
library(taxastand)
data(filmy_taxonomy)

dir <- tempfile("bench")
dir.create(dir)

# Best-of-n elapsed seconds for a zero-argument function
timeit <- function(f, reps = 2) {
  min(vapply(
    seq_len(reps),
    function(i) system.time(f())[["elapsed"]], numeric(1)
  ))
}

# Run a taxon-tools command in the Docker image, mounting `dir` at /data
docker_run <- function(args) {
  system2("docker", c("run", "--rm", "-v", paste0(dir, ":/data"), image, args),
    stdout = TRUE, stderr = FALSE
  )
}

base <- unique(filmy_taxonomy$scientificName)
base <- base[!is.na(base) & nzchar(base)]
length(base)
```

## Parsing

```{r parse-bench, eval = has_docker}
parse_rows <- lapply(c(1, 5, 10), function(mult) {
  names_vec <- rep(base, mult)
  records <- paste0("n", seq_along(names_vec), "|", names_vec)
  writeLines(records, file.path(dir, "p.txt"), useBytes = TRUE)
  n <- length(records)

  r_t <- timeit(function() taxastand:::tt_parsenames(records))
  d_t <- timeit(function() docker_run(c("parsenames", "/data/p.txt")))

  data.frame(
    `Names parsed` = n,
    `pure R (s)` = round(r_t, 2),
    `taxon-tools (s)` = round(d_t, 2),
    `R names/s` = round(n / r_t),
    `Speed-up` = sprintf("%.1fx", d_t / r_t),
    check.names = FALSE
  )
})
parse_tbl <- do.call(rbind, parse_rows)
```

```{r parse-table, eval = has_docker}
knitr::kable(parse_tbl, align = "r")
```

## Matching

The reference (the whole taxonomy) and a query set (misspellings and
author-dropped variants of a sample) are parsed up front, so that only the
matching step is timed.

```{r match-bench, eval = has_docker}
parse_lines <- function(x, prefix) {
  lines <- taxastand:::tt_parsenames(paste0(prefix, seq_along(x), "|", x))
  rest <- sub("^[^|]*\\|", "", lines)
  lines[gsub("\\|", "", rest) != ""] # drop unparseable names
}

ref_lines <- parse_lines(base, "r")

set.seed(1)
samp <- sample(base, 300)
mis <- vapply(samp, function(s) {
  cs <- strsplit(s, "")[[1]]
  pos <- which(grepl("[a-z]", cs))
  if (length(pos) > 3) cs[pos[length(pos) %/% 2]] <- "x"
  paste(cs, collapse = "")
}, character(1))
query <- unique(c(samp, mis, sub(" [A-Z(].*$", "", samp)))
query_lines <- parse_lines(query, "q")

writeLines(ref_lines, file.path(dir, "ref.txt"), useBytes = TRUE)
writeLines(query_lines, file.path(dir, "query.txt"), useBytes = TRUE)

match_rows <- lapply(c(5, 10), function(e) {
  r_t <- timeit(function() {
    taxastand:::tt_matchnames(query_lines, ref_lines, max_dist = e)
  })
  d_t <- timeit(function() {
    docker_run(c(
      "matchnames", "-a", "/data/query.txt", "-b", "/data/ref.txt",
      "-o", "/data/out.txt", "-e", e, "-F"
    ))
  })

  data.frame(
    `max_dist` = e,
    `pure R (s)` = round(r_t, 2),
    `taxon-tools (s)` = round(d_t, 2),
    `Speed-up` = sprintf("%.1fx", d_t / r_t),
    check.names = FALSE
  )
})
match_tbl <- do.call(rbind, match_rows)
```

```{r match-table, eval = has_docker}
cat(sprintf(
  "%d queries against %d references\n",
  length(query_lines), length(ref_lines)
))
knitr::kable(match_tbl, align = "r")
```

```{r no-docker-note, eval = !has_docker, echo = FALSE, results = "asis"}
cat(
  "> **Docker (or the `", image, "` image) was not available when this ",
  "vignette was built, so the benchmark was not run.** Install Docker and ",
  "pull the image, then re-render this vignette to see live results.\n",
  sep = ""
)
```

## Takeaways

```{r takeaway-numbers, eval = has_docker, echo = FALSE, results = "asis"}
p1 <- parse_tbl[parse_tbl$`Names parsed` == length(base), ]
m10 <- match_tbl[match_tbl$max_dist == 10, ]
cat(sprintf(
  paste0(
    "On this machine, parsing %s names took %.2fs in pure R vs %.2fs via ",
    "taxon-tools (%s), and matching at the default `max_dist = 10` was ",
    "%s faster.\n"
  ),
  format(length(base), big.mark = ","),
  p1$`pure R (s)`, p1$`taxon-tools (s)`, p1$`Speed-up`, m10$`Speed-up`
))
```

- **Parsing** is several times faster in pure R. Both implementations have
  near-constant per-name cost; the gap is largest for small inputs, where
  Docker's container start-up dominates, but pure R stays well ahead even at
  tens of thousands of names.

- **Matching** is where the difference is dramatic, and it *grows* with
  `max_dist`. The original tool fuzzy-matches with gawk's `amatch()` (TRE
  approximate-regex matching), whose cost climbs steeply as the allowed edit
  distance increases. The pure-R engine uses base R's vectorised `adist()`
  restricted to within-genus candidates, which barely moves from `max_dist` 5
  to 10.

- Beyond raw speed, the pure-R implementation removes the need to install
  `taxon-tools` or run Docker at all, while producing identical results.

These numbers depend on hardware, dataset size, and the proportion of names that
require fuzzy matching; re-render this vignette to measure them on your machine.

```{r session, eval = has_docker}
sessionInfo()
```
